Factors Influencing Community Garden Growth: Analyzing Trends in PollinateTO Funding

### 2024-12-15 paper/paper.qmd
Update paper

---
title: "Key Drivers of Community Garden Growth: Analyzing Trends in PollinateTO Funding"
subtitle: "A Bayesian Approach to Understanding the Impact of Time, Type, and Location on Garden Size and Number in Toronto's Neighborhoods from 2020 to 2023."
author: 
  - Aliza Abbas Mithwani
thanks: "Code and data are available at: [https://github.com/alizamithwani/PollinateTO.git]"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format:
  pdf:
    toc: true
number-sections: true
header-includes:
  - \usepackage{pdflscape}
bibliography: references.bib
---

@@

#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(arrow)
library(rstanarm)
library(patchwork)
library(tinytable)
library(kableExtra)
library(broom.mixed)

@@

## Overview
Urban biodiversity is increasingly recognized as vital to maintaining healthy ecosystems, especially in the face of rapid urbanization. Among the most important urban biodiversity goals is the protection of pollinators, whose decline poses significant risks to food security, agricultural productivity, and ecological stability. In response to this, many cities have launched initiatives to create pollinator-friendly habitats within urban spaces. Toronto’s PollinateTO program is one such initiative, aiming to enhance biodiversity by funding community-led pollinator gardens. Since its launch in 2019, the program has supported over 190 projects, resulting in nearly 500 gardens and approximately 25,500 square meters of pollinator habitat.

Despite the program's success in expanding green spaces, the allocation of funding across these projects remains an understudied area. Questions about which types of gardens or neighborhoods are more likely to receive funding, and the factors influencing these decisions, are crucial for optimizing the program’s impact. The objective of this paper is to examine these allocation patterns, specifically whether garden characteristics such as type (e.g., rain gardens, boulevard gardens) or location (e.g., Neighbourhood Improvement Areas) influence the likelihood of receiving funding and the size of the funding allocation.

In particular, this paper addresses the gap in understanding how urban biodiversity funding decisions are made, with a focus on the specific factors that contribute to higher funding likelihoods and amounts. The estimand in this analysis is whether particular garden features, community engagement strategies, or geographic locations are associated with receiving more substantial funding allocations from the PollinateTO program. By identifying these key factors, this research provides important insights into the effectiveness and equity of funding distributions for urban biodiversity projects.

Our findings indicate that garden type, community engagement, and the location of the project all play significant roles in funding decisions. Specifically, we find that gardens in underserved neighborhoods and those emphasizing community involvement are more likely to receive higher funding allocations. These results offer actionable insights for optimizing resource distribution in urban biodiversity programs and aligning ecological goals with social equity considerations.

The remainder of the paper is structured as follows: @sec-data describes the data used in the analysis, including its sources, key features, and considerations regarding measurement and visualization. Section 3 outlines the methodology, detailing the data visualization techniques employed and the justification for using a Bayesian logistic regression model to analyze the relationship between garden characteristics and funding outcomes. Section 4 presents the results, including model validation and key findings. Section 5 offers a discussion of the results, focusing on their implications through key points, followed by an examination of the study's limitations and suggested future steps for further exploration. Section 6 concludes the paper. The Appendix provides supplementary materials, including raw data previews, data feature definitions, data visualizations, and technical model details such as posterior predictive checks, MCMC convergence diagnostics, and additional figures.

# Data {#sec-data}
## Software Packages
This analysis utilizes the R programming language [@citeR], with the `rstanarm` [@rstanarm] and `tidyverse` [@tidyverse] packages for statistical modeling and data manipulation. Our data comes from the PollinateTO Primary Project Garden Locations dataset [@PTdata], which is hosted on Toronto's Open Data Portal. Following the approach outlined in @tellingstories, we leverage these tools to explore patterns and trends in the allocation of PollinateTO funding.

## Data Source
The primary source of data for this study is the PollinateTO Primary Project Garden Locations dataset, which contains detailed information on community-led garden projects funded by PollinateTO. The dataset includes key features such as project locations, garden types, funding years, and other relevant information that enables an in-depth analysis of funding patterns and their spatial and socio-economic implications. This dataset is publicly accessible through Toronto's Open Data Portal and is a vital resource for studying the impact and effectiveness of PollinateTO in promoting urban biodiversity.

## Data Features
The dataset contains several key variables that are critical for understanding the distribution and characteristics of funded garden projects. Below, we describe each variable and its relevance to the analysis:

- YEAR_FUNDED: The year each project received funding from PollinateTO. This variable is crucial for understanding temporal trends in the program’s funding allocation. It also allows us to analyze shifts in funding priorities over time.
- GARDEN_TYPE: This variable categorizes the gardens by their ecological functions (e.g., rain gardens, food gardens, pollinator gardens). These types reflect the intended ecological role of each garden and are important for understanding the program's impact on biodiversity and the environment.
- NIA_OR_EN: This classification indicates whether the project is located in a Neighbourhood Improvement Area (NIA), Emerging Neighbourhood (EN), or another area. These labels are derived from Toronto's planning data and allow for an exploration of whether funding is disproportionately allocated to areas with different socio-economic characteristics, enabling us to analyze equity in the distribution of funds.
- WARD_NAME: The specific ward in Toronto where the garden project is located. This geographic indicator helps us assess whether certain neighborhoods receive more funding based on local government priorities or demographic characteristics.
- IS_INDIGENOUS_GARDEN: A binary variable that identifies whether the project is Indigenous-led, based on self-identification in the funding application. This is a key variable for exploring whether the program is aligned with Toronto’s commitments to Indigenous communities and ecological knowledge.
PROJECT_NAME: The name assigned to each garden project. While this variable is useful for identification, it is not directly involved in the analysis but helps in linking data points to specific projects.
- LATITUDE and LONGITUDE: These geographic coordinates provide spatial context for each garden project. While these data points help in mapping and visualizing the geographic distribution of funded gardens, their primary role is in understanding spatial patterns and localizing project locations.

@@

## Constructed Variables
In this study, variable is_large_garden was constructed. This binary variable was constructed to classify gardens as "large" or "small" based on their estimated garden size. Specifically, gardens with an estimated garden size greater than the mean size of all gardens in the dataset were coded as 1 (large garden), while those with a size less than or equal to the mean were coded as 0 (small garden). The mean garden size was calculated using the estimated_garden_size variable, excluding any missing values. This classification was chosen to capture differences in garden size relative to the overall distribution and to investigate whether larger gardens are more likely to receive funding.

@@

Granularity: Some ecological measures, such as specific plant species or detailed environmental outcomes, are not captured in the dataset. This restricts the scope of ecological impact analysis, though the dataset’s focus on garden characteristics and funding allocation provides valuable insights.

@@

## Methodology
The PollinateTO dataset collects detailed information on community-led garden projects funded by Toronto’s PollinateTO initiative. Data is sourced from the city’s Open Data portal and includes variables like garden type, location, funding year, and size, enabling analysis of trends in urban biodiversity projects. The dataset uses standardized attributes and is updated annually, making it a valuable tool for evaluating the initiative’s impact across various Toronto neighborhoods, though it may be subject to self-reporting biases and geographical limitations in precision.
\newpage

@@

# Use the log-transformed column in the dotplot
ggplot(PT_analysis_data, aes(x = nia_or_en, y = log_garden_size, fill = nia_or_en)) +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.8) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Log-Transformed Garden Sizes by Neighborhood Type",
    x = "Neighborhood Type (NIA or EN)",
    y = "Log of Estimated Garden Size",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
# Model
In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project. We are interested in modeling whether a garden is large (binary outcome) as a function of various factors such as the garden type, year funded, neighborhood type, indigenous garden status, and ward name. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

\begin{align} 
y_i|\pi_i & \sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha 
    + \beta_1 \times \mbox{garden type}_i 
    + \beta_2 \times \mbox{year funded}_i \notag \\
    &\quad + \beta_3 \times \mbox{nia or en}_i 
    + \beta_4 \times \mbox{is indigenous garden}_i 
    + \beta_5 \times \mbox{ward name}_i \notag \\
    &\quad + \beta_6 \times (\mbox{garden type}_i \times \mbox{nia or en}_i) \\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\beta_4 &\sim \mbox{Normal}(0, 2.5) \\
\beta_5 &\sim \mbox{Normal}(0, 2.5)
\end{align}

@@

### Model justification

@@

The choice of the Bayesian approach in this context also allows for the flexible inclusion of multiple categorical predictors, which is crucial for this study, as it aims to assess the relationship between categorical factors like garden type, ward names, and the interaction between garden type and nia or en with a quantitative outcome. The model's use of informative priors helps regularize the estimates, particularly when dealing with categorical predictors that may have complex relationships with the outcome variable. Additionally, the Bayesian methodology provides the opportunity to assess model diagnostics comprehensively, ensuring that the model fits well with the data and that the parameters are estimated with high precision. Overall, this model is well-suited to capture the complexities of the data while offering interpretability and flexibility through its probabilistic nature.

# Results

Our results are summarized in @tbl-modeltable. The Bayesian logistic regression model reveals several important relationships between garden characteristics and the likelihood of a garden being classified as large. Among garden types, "Multiple Property Garden" stands out with a substantial positive coefficient (14.94), suggesting that such gardens are strongly associated with larger sizes. Conversely, "Boulevard Garden, Rain Garden" exhibits a notable negative coefficient (-13.77), indicating a reduced likelihood of being large. Other garden types, such as "Communal Garden" and "School Learning Garden," have smaller coefficients, implying a weaker or negligible relationship with garden size. This variability in the influence of garden type highlights the importance of understanding project-specific features when predicting garden outcomes.

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modeltable
#| warning: false
#| tbl-cap: "Explanatory model of whether a garden is large based on certain factors: Indigenous-led indicator, Year Funded, Neighborhood, Garden type and Ward Names"

@@

# Render the table using kableExtra
model_table %>%
  kbl() %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  landscape()
```
@fig-model-coefficients visualizes the estimated effects of various predictors on the likelihood of a garden being classified as "large" in the PollinateTO initiative. Each point represents the coefficient estimate for a predictor, indicating the direction and strength of its relationship with garden size. Predictors with positive coefficients increase the probability of a garden being large, while those with negative coefficients decrease it. The horizontal bars show the 95% credible intervals, capturing the uncertainty in the estimates. Predictors whose credible intervals do not cross the dashed red line at zero are likely to have significant effects. This plot highlights which variables, such as garden type, neighborhood classification, or year funded, are most influential in determining garden size, providing valuable insights into the factors driving outcomes in the program.

@@

# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(rstanarm)

@@

\newpage
# Discussion

## Unearthing Patterns in Garden Allocation {#sec-first-point}
This paper investigates the factors influencing the size of gardens funded under Toronto’s PollinateTO initiative using a Bayesian logistic regression model. Specifically, it explores how garden size (a binary outcome of “large” vs. “not large”) is associated with variables such as garden type, year funded, neighborhood classification (Emerging Neighborhoods or Neighborhood Improvement Areas), Indigenous-led status, and ward-level geographic differences. By leveraging this model, the analysis identifies patterns and disparities in garden allocation, providing insights into which factors are most predictive of larger gardens. The approach incorporates interaction effects to capture nuanced relationships between garden types and neighborhood characteristics.

The inclusion of credible intervals from the Bayesian model ensures a robust interpretation of uncertainty in coefficient estimates. Additionally, key goodness-of-fit measures (e.g., R², WAIC) are evaluated, demonstrating the model’s ability to explain nearly half of the variation in garden size. These findings offer a data-driven lens through which urban sustainability programs like PollinateTO can be assessed and improved.

@@

One of the central insights from this analysis is the existence of disparities in garden size across neighborhoods. Specifically, gardens in areas classified as “None” (neither EN nor NIA) are significantly more likely to be large, while gardens in NIAs face challenges in achieving similar sizes. This suggests that systemic factors, such as socioeconomic inequalities, land availability, and access to resources, influence the outcomes of urban greening initiatives.

This insight highlights a broader inequity in urban sustainability programs: neighborhoods with fewer resources or systemic disadvantages may struggle to achieve comparable environmental benefits. This finding emphasizes the importance of embedding equity-focused considerations into grant allocation processes. Addressing these disparities can improve not only the environmental impact of such initiatives but also their social relevance and inclusivity.

@@

Indigenous-led projects are less likely to result in large gardens, pointing to potential barriers that Indigenous communities face in participating fully in urban greening initiatives. This finding underscores the need to reevaluate the criteria for “success” in such programs. Indigenous-led projects may prioritize cultural and ecological goals—such as the preservation of native plant species or community education—over garden size.

This raises an important consideration for future urban sustainability programs: qualitative impacts, such as community engagement or cultural significance, may be equally (if not more) important than quantitative measures like garden size. By failing to account for these dimensions, existing frameworks risk marginalizing projects that offer critical social and cultural benefits. This insight encourages a shift in how success is defined and measured in urban greening efforts.

# Weaknesses and next steps
I will address some of the weaknesses and limitations of the current analysis, focusing on areas where the methodology or data could be improved. Following this, I will explore potential next steps, outlining opportunities for future research and policy development that build on these findings to create more equitable and effective urban sustainability initiatives. This section aims to critically reflect on the study while charting a path forward for both academic inquiry and practical applications.

## Challenges and Limitations in Modeling Garden Outcomes
While the model provides valuable insights, it is not without limitations. First, the binary categorization of garden size as “large” vs. “not large” may oversimplify the complexity of garden outcomes. A more nuanced approach, such as modeling size as a continuous variable or including qualitative measures of success, might provide richer insights.

Second, the dataset is limited to gardens funded by PollinateTO, which may not fully represent the broader urban gardening landscape in Toronto. Selection bias could influence the generalizability of these findings to other urban greening initiatives. Additionally, the potential for omitted variable bias exists, as factors like community engagement levels or prior gardening experience were not included in the analysis but may significantly impact garden outcomes.

Finally, while interaction terms capture some complexity, they add interpretive challenges, as not all combinations of garden types and neighborhoods had sufficient data to provide precise estimates. Future work could address this by gathering larger, more diverse datasets or employing hierarchical models to account for varying group sizes.

## Opportunities for Future Research and Policy Improvements
Several open questions remain. One key area for future research is the inclusion of community-level variables, such as demographic characteristics, access to volunteer networks, and proximity to green spaces. These factors may help explain why certain neighborhoods are more likely to host large gardens.

Additionally, further work is needed to explore the qualitative outcomes of urban gardens, particularly for Indigenous-led projects. Incorporating interviews, case studies, or community feedback could provide a deeper understanding of how these gardens contribute to cultural preservation and community well-being.

From a methodological perspective, future studies could explore alternative modeling techniques, such as hierarchical Bayesian models, to better account for ward-level variability and cross-neighborhood interactions. Expanding the dataset to include other urban greening initiatives or longitudinal data could also provide insights into how garden outcomes evolve over time.

Lastly, there is a need to test the effectiveness of policy interventions informed by these findings. For instance, targeted funding for NIAs or revised evaluation criteria for Indigenous-led projects could be piloted and evaluated to determine their impact on garden size and community engagement. This iterative process of evidence-based policy design and evaluation can help create more equitable and effective urban sustainability programs.

>>>>

---
title: "Factors Influencing Community Garden Growth: Analyzing Trends in PollinateTO Funding"
subtitle: "A Bayesian Approach to Understanding the Impact of Time, Type, and Location on Garden Size and Number in Toronto's Neighborhoods from 2020 to 2023."
author: 
  - Aliza Abbas Mithwani
thanks: "Code and data are available at: [https://github.com/alizamithwani/PollinateTO.git]"
date: today
date-format: long
abstract: "This paper investigates the allocation patterns of grants for pollinator gardens funded by the PollinateTO initiative in Toronto, focusing on how garden characteristics, neighborhood type, and community engagement influence funding decisions. A Bayesian logistic regression model reveals that gardens located in Neighbourhood Improvement Areas (NIA) or Emerging Neighborhoods (EN) are more likely to receive higher funding, especially for specific garden types like Multi-Residential Gardens and Multiple Property Gardens. Additionally, Indigenous-led gardens and those located in certain wards (e.g., Scarborough Centre) tend to receive larger grants. This research contributes to optimizing the allocation of urban biodiversity funding, ensuring equitable support for community-driven ecological projects."
format:
  pdf:
    toc: true
number-sections: true
header-includes:
  - \usepackage{pdflscape}
bibliography: references.bib
---

@@

#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(arrow)
library(rstanarm)
library(patchwork)
library(kableExtra)
library(broom.mixed)

@@

Urban biodiversity is recognized as vital to maintaining healthy ecosystems, particularly in the context of rapid urbanization. A key goal is the protection of pollinators, whose decline presents risks to food security, agricultural productivity, and ecological stability. In response to this, many cities have launched initiatives to create pollinator-friendly habitats within urban areas. Toronto’s PollinateTO program is one such initiative, aiming to enhance biodiversity by funding community-led pollinator gardens. Since its launch in 2019, the program has supported over 190 projects, resulting in nearly 500 gardens and approximately 25,500 square meters of pollinator habitat.

Despite the program's success in expanding green spaces, the allocation of funding across these projects remains an under-explored area. Questions about which types of gardens or neighborhoods are more likely to receive funding, and the factors influencing these decisions, are important for optimizing the program’s impact. The objective of this paper is to examine these allocation patterns, specifically whether garden characteristics such as type (e.g., rain gardens, boulevard gardens) or location (e.g., Neighbourhood Improvement Areas) influence the likelihood of receiving funding and the size of the funding allocation.

This paper addresses the lack of understanding how urban biodiversity funding decisions are made, with a focus on the factors that contribute to higher funding likelihoods and amounts. The estimand in this analysis is whether particular garden features, community engagement strategies, or geographic locations are associated with receiving larger funding allocations from the PollinateTO program. By identifying these factors, this research provides important findings related to the effectiveness and equity of funding distributions for urban biodiversity projects.

Our findings indicate that garden type, community engagement, and the location of the project all play significant roles in funding decisions. Specifically, we find that gardens in underserved neighborhoods and those emphasizing community involvement are more likely to receive larger funding allocations. These results offer actionable recommendations for optimizing resource distribution in urban biodiversity programs and aligning ecological goals with social equity considerations.

The remainder of the paper is structured as follows: @sec-data describes the data used in the analysis, including its sources, key features, and considerations regarding measurement and visualization. @sec-model outlines the methodology, detailing the data visualization techniques employed and the justification for using a Bayesian logistic regression model to analyze the relationship between garden characteristics and funding outcomes. @sec-results presents the results, including model validation and key findings. @sec-discussion offers a discussion of the results, focusing on their implications through key points, followed by an examination of the study's limitations and suggested future steps for further analysis. @sec-finals concludes the paper. The {-} provides supplementary materials, including additional model details related to cleaning and simulation, a section about survey, sampling, and observational data in the context of this paper, and technical model details such as posterior predictive checks and MCMC convergence diagnostics.

# Data {#sec-data}
## Software Packages
This analysis utilizes the R programming language [@citeR], with the `rstanarm` [@rstanarm], `tidyverse` [@tidyverse], `opendatatoronto` [@opendatatoronto], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `knitr` [@knitr], `arrow` [@arrow], `patchwork` [@patchwork], `kableExtra` [@kableExtra], and `broom.mixed` [@broommixed] packages for statistical modeling and data manipulation. Our data comes from the PollinateTO Primary Project Garden Locations dataset [@PTdata], which is hosted on Toronto's Open Data Portal. Following the approach outlined in @tellingstories, we use these tools to explore patterns and trends in the allocation of PollinateTO funding.

## Data Source
The primary source of data for this study is the PollinateTO Primary Project Garden Locations dataset [@PTdata], which contains detailed information on community-led garden projects funded by PollinateTO. The dataset includes key features such as project locations, garden types, funding years, and other relevant information that enables an in-depth analysis of funding patterns and their spatial and socio-economic implications. This dataset is publicly accessible through Toronto's Open Data Portal and is a vital resource for studying the impact and effectiveness of PollinateTO in promoting urban biodiversity.

## Data Features
The dataset contains several key variables that are important for understanding the distribution and characteristics of funded garden projects. Below, we describe each variable and its relevance to the analysis:

- YEAR_FUNDED: The year each project received funding from PollinateTO. This variable is important for understanding temporal trends in the program’s funding allocation. It also allows us to analyze shifts in funding priorities over time.
- GARDEN_TYPE: This variable categorizes the gardens by their ecological functions (e.g., rain gardens, food gardens, pollinator gardens). These types reflect the intended ecological role of each garden and are important for understanding the program's impact on biodiversity and the environment.
- NIA_OR_EN: This classification indicates whether the project is located in a Neighbourhood Improvement Area (NIA), Emerging Neighbourhood (EN), or another area. These labels are derived from Toronto's planning data and allow for a search of whether funding is disproportionately allocated to areas with different socio-economic characteristics, enabling us to analyze equity in the distribution of funds.
- WARD_NAME: The specific ward in Toronto where the garden project is located. This geographic indicator helps us assess whether certain neighborhoods receive more funding based on local government priorities or demographic characteristics.
- IS_INDIGENOUS_GARDEN: A binary variable that identifies whether the project is Indigenous-led, based on self-identification in the funding application. This is a key variable for exploring whether the program is aligned with Toronto’s commitments to Indigenous communities and ecological knowledge.
- PROJECT_NAME: The name assigned to each garden project. While this variable is useful for identification, it is not directly involved in the analysis but helps in linking data points to specific projects.
- LATITUDE and LONGITUDE: These geographic coordinates provide spatial context for each garden project. While these data points help in mapping and visualizing the geographic distribution of funded gardens, their primary role is in understanding spatial patterns and localizing project locations.

@@

## Constructed Variables
In this study, the variable is_large_garden was constructed. This binary variable was created to classify gardens as "large" or "small" based on their estimated garden size. Specifically, gardens with an estimated garden size greater than the mean size of all gardens in the dataset were coded as 1 (large garden), while those with a size less than or equal to the mean were coded as 0 (small garden). The mean garden size was calculated using the estimated_garden_size variable, excluding any missing values. This classification was chosen to capture differences in garden size relative to the overall distribution and to investigate whether larger gardens are more likely to receive funding.

@@

Granularity: Some ecological measures, such as specific plant species or detailed environmental outcomes, are not captured in the dataset. This restricts the scope of ecological impact analysis, though the dataset’s focus on garden characteristics and funding allocation provides important information.

@@

## Methodology
The PollinateTO dataset collects detailed information on community-led garden projects funded by Toronto’s PollinateTO initiative. Data is sourced from the city’s Open Data portal and includes variables like garden type, location, funding year, and size, enabling analysis of trends in urban biodiversity projects. The dataset uses standardized attributes and is updated annually, making it a useful tool for evaluating the initiative’s impact across various Toronto neighborhoods, though it may be subject to self-reporting biases and geographical limitations in precision.
\newpage

@@

# Use the log-transformed column in the dotplot
ggplot(PT_analysis_data, aes(x = nia_or_en, y = log_garden_size, fill = nia_or_en)) +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.8) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Log-Transformed Garden Sizes by Neighborhood Type",
    x = "Neighborhood Type (NIA or EN)",
    y = "Log of Estimated Garden Size",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
# Model {#sec-model}
In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project. We are interested in modeling whether a garden is large (binary outcome) as a function of various factors such as the garden type, year funded, neighborhood type, indigenous garden status, and ward name. Background details and diagnostics are included in [Appendix - @sec-model-details].

@@

\begin{align} 
y_i|\pi_i & \sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha 
    + \beta_1 \times \mbox{garden type}_i 
    + \beta_2 \times \mbox{year funded}_i \notag \\
    &\quad + \beta_3 \times \mbox{nia or en}_i 
    + \beta_4 \times \mbox{is indigenous garden}_i 
    + \beta_5 \times \mbox{ward name}_i \notag \\
    &\quad + \beta_6 \times (\mbox{garden type}_i \times \mbox{nia or en}_i) \\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\beta_4 &\sim \mbox{Normal}(0, 2.5) \\
\beta_5 &\sim \mbox{Normal}(0, 2.5) \\
\beta_6 &\sim \mbox{Normal}(0, 2.5)
\end{align}

@@

## Model justification

@@

The choice of the Bayesian approach in this context also allows for the flexible inclusion of multiple categorical predictors, which is important for this study, as it aims to assess the relationship between categorical factors like garden type, ward names, and the interaction between garden type and nia or en with a quantitative outcome. The model's use of informative priors helps regularize the estimates, particularly when dealing with categorical predictors that may have complex relationships with the outcome variable. Additionally, the Bayesian methodology provides the opportunity to assess model diagnostics, ensuring that the model fits well with the data and that the parameters are estimated with high precision. Overall, this model is well-suited to capture the complexities of the data while offering interpretability and flexibility through its probabilistic nature.

# Results {#sec-results}

Our results are summarized in @tbl-modeltable. The Bayesian logistic regression model shows several important relationships between garden characteristics and the likelihood of a garden being classified as large. Among garden types, "Multiple Property Garden" stands out with a substantial positive coefficient (14.94), suggesting that such gardens are strongly associated with larger sizes. Conversely, "Boulevard Garden, Rain Garden" exhibits a notable negative coefficient (-13.77), indicating a reduced likelihood of being large. Other garden types, such as "Communal Garden" and "School Learning Garden," have smaller coefficients, implying a weaker or negligible relationship with garden size. This variability in the influence of garden type highlights the importance of understanding project-specific features when predicting garden outcomes.

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modeltable
#| warning: false
#| tbl-cap: "Explanatory model of whether a garden is large based on certain factors: Indigenous-led indicator, Year Funded, Neighborhood, Garden type and Ward Names (zoom in for a clearer view)"

@@

# Render the table using kableExtra
model_table %>%
  kbl() %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  landscape() %>%
  column_spec(1, width = "1cm") %>%
  kable_styling(font_size = 3)  # Reduce the font size to make it smaller
```
@fig-model-coefficients visualizes the estimated effects of various predictors on the likelihood of a garden being classified as "large" in the PollinateTO initiative. Each point represents the coefficient estimate for a predictor, indicating the direction and strength of its relationship with garden size. Predictors with positive coefficients increase the probability of a garden being large, while those with negative coefficients decrease it. The horizontal bars show the 95% credible intervals, capturing the uncertainty in the estimates. Predictors whose credible intervals do not cross the dashed red line at zero are likely to have significant effects. This plot highlights which variables, such as garden type, neighborhood classification, or year funded, are most influential in determining garden size, providing understanding of the factors driving outcomes in the program.

@@

@@

\newpage
# Discussion {#sec-discussion}

## Unearthing Patterns in Garden Allocation {#sec-first-point}

This paper investigates the factors influencing the size of gardens funded under Toronto’s PollinateTO initiative using a Bayesian logistic regression model. Specifically, it explores how garden size, defined as a binary outcome (large vs. not large), is associated with several key predictors: garden type, year funded, neighborhood classification (Emerging Neighborhoods or Neighborhood Improvement Areas), Indigenous-led status, and ward-level geographic differences. The analysis shows several noteworthy patterns and disparities in the allocation of larger gardens. By leveraging this model, the study identifies which factors have the most significant influence on whether a garden is classified as large. This approach is enhanced by incorporating interaction effects, which provide a deeper understanding of the relationships between garden types and neighborhood characteristics.

The Bayesian logistic regression framework offers a probabilistic interpretation of these relationships, allowing for a more complex view of uncertainty in parameter estimates. By incorporating credible intervals, the model quantifies the uncertainty around the estimated coefficients, providing a more robust and transparent analysis compared to traditional frequentist methods. In addition to providing understanding of the predictive factors of garden size, this paper also assesses model fit through key measures such as R² and WAIC. These measures indicate that the model explains a significant portion of the variation in garden size, with nearly half of the variation accounted for by the chosen predictors.

The implications of these findings are significant for urban sustainability programs like PollinateTO. By highlighting the key drivers of garden size, this research offers an analytical lens through which policymakers can refine the allocation process for funding community gardens. If the goal of such initiatives is to foster environmental sustainability and equity, then understanding which factors contribute to larger gardens is important. For instance, urban greening efforts that are situated in more resource-rich neighborhoods or that prioritize certain types of gardens may need to be adjusted to address inequities and ensure that all communities have access to the same opportunities for greening and sustainability.

Additionally, the interaction terms between garden type and neighborhood classification are particularly important. They show that certain garden types (e.g., vegetable or pollinator gardens) tend to be more successful in certain types of neighborhoods, which may point to localized resources or specific neighborhood priorities. This finding suggests that urban greening programs should consider tailoring their strategies based on local needs and resources. Programs that adopt a more context-sensitive approach, considering both neighborhood characteristics and garden types, are likely to see more equitable and effective outcomes.

@@

A major finding from this analysis is the evident disparities in garden size across different neighborhoods. Specifically, gardens in areas classified as “None” (i.e., neighborhoods neither in the Emerging Neighborhoods nor in the Neighborhood Improvement Areas) are significantly more likely to be classified as large, while gardens located in Neighborhood Improvement Areas (NIAs) are less likely to achieve the same size. This finding suggests that neighborhoods with fewer systemic challenges, such as better access to resources, land availability, and socioeconomic stability, are more likely to develop large, successful urban gardens. Conversely, communities in NIAs, which often face challenges like poverty, limited access to green space, and lower levels of community investment, struggle to achieve similar outcomes.

This disparity underscores a broader issue in urban sustainability programs: systemic factors, such as socioeconomic inequalities, land availability, and access to resources, significantly affect the success of these initiatives. These inequalities are often compounded by historical disinvestment in marginalized neighborhoods, which can lead to lower levels of community engagement and reduced capacity for sustaining urban greening efforts. The difference in garden size between these neighborhoods points to the need for equity-focused policies in urban greening initiatives.

The implication is clear: sustainability programs must address neighborhood-level disparities in access to resources and opportunities. While it is important to continue fostering urban sustainability, it is equally important to ensure that these efforts benefit all communities, particularly those that have historically been underserved. Programs like PollinateTO should consider revising their funding allocation strategies to prioritize neighborhoods with fewer resources or greater systemic disadvantages. This could involve providing additional technical assistance, offering capacity-building programs for community members, or targeting more funding toward NIAs to ensure these communities can achieve similar environmental and social benefits. By doing so, urban sustainability programs can more effectively address environmental justice concerns, ensuring that all communities have the opportunity to benefit from urban greening.

Moreover, this finding points to a need for broader systemic change. To truly address inequities in urban gardening, policymakers must also focus on improving the underlying conditions in marginalized neighborhoods. This could include initiatives to improve access to affordable housing, better public transportation, and community infrastructure. Without such systemic changes, urban greening programs may fail to achieve their full potential in creating equitable and sustainable urban environments.

@@

Indigenous-led projects in urban greening initiatives, including those funded by PollinateTO, are less likely to result in large gardens compared to other projects. This finding raises important questions about the criteria used to evaluate the success of these projects. Indigenous communities may prioritize different values in their urban greening efforts, focusing on goals such as the preservation of native plant species, the restoration of traditional ecological practices, and the fostering of community cohesion, rather than on maximizing garden size. As a result, the existing framework for evaluating the success of urban gardens, which focuses primarily on the quantitative measure of size, may not fully capture the diverse goals and impacts of Indigenous-led initiatives.

This finding calls for a reevaluation of how success is defined and measured in urban sustainability programs. While garden size is a useful indicator of certain types of success, it may not adequately capture the full range of outcomes that Indigenous-led projects aim to achieve. For example, many Indigenous communities prioritize cultural, social, and ecological sustainability over the mere expansion of garden size. Gardens that are smaller but culturally significant or that provide vital ecosystem services may be just as important, if not more so, than larger gardens that lack these dimensions.

The need to redefine success also speaks to a broader issue in urban sustainability programs: the tendency to prioritize Western, quantitative metrics of success, such as garden size, over qualitative outcomes that may be more meaningful to Indigenous communities. This discrepancy often leads to the marginalization of projects that offer important cultural and social benefits but do not conform to conventional measures of success. By failing to account for these dimensions, urban sustainability frameworks risk overlooking initiatives that are culturally relevant and community-driven.

This finding suggests a shift toward a more inclusive and holistic approach to evaluating urban greening projects. Success should not be measured solely by garden size, but rather by the broader social, cultural, and environmental impacts that these gardens have on the community. Future iterations of urban greening programs should consider the inclusion of qualitative measures, such as community engagement, cultural preservation, and ecological restoration, alongside traditional quantitative metrics. This would ensure that Indigenous-led projects are evaluated on their own terms and that they receive the recognition and support they deserve.

# Weaknesses and next steps {#sec-finals}

While this analysis provides important findings, it is important to acknowledge the limitations of the current study and suggest areas for improvement in both the methodology and data collection. Addressing these weaknesses will help refine the findings and lead to more robust and equitable urban sustainability programs.

## Challenges and Limitations in Modeling Garden Outcomes

The model employed in this study provides a useful understanding of the factors influencing garden size, but it has several limitations. One of the primary weaknesses is the binary categorization of garden size into “large” vs. “not large.” This binary approach oversimplifies the complexity of garden outcomes, as garden size is a continuous variable that may better reflect the true diversity of urban gardens. Future research could model garden size as a continuous variable, allowing for a deeper understanding of how factors such as neighborhood characteristics and garden type influence garden size. Additionally, including qualitative measures of success, such as the impact of the garden on local biodiversity or community engagement, would enrich the analysis and provide a more thorough understanding of garden outcomes.

Another limitation is the focus on gardens funded by the PollinateTO initiative. While this dataset provides important findings, it may not fully represent the broader urban gardening landscape in Toronto. The sample is limited to one specific initiative, and the findings may not be generalizable to other urban greening programs or gardening projects outside of Toronto. Moreover, the potential for selection bias exists, as the gardens included in the dataset are those that were selected for funding, and these may not be representative of all gardens in the city. Future studies could broaden the scope to include data from other urban greening initiatives or include non-funded gardens to improve the generalizability of the findings.
